Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2024-12-02 responseToReviews.md

Adaptation effects.  Adapted contrast response functions have been measured by Nelson et al., (EEG Journal, 1984) by comparing CRFs measured by sweeping contrast from low to high vs high to low. They found that responses at low contrast were smaller in the “down” sweeps and attributed it to differences in prior adaptation.

> we could not find this reference. We did find an excellent (and sometimes humorous) review of SSVEPs by Nelson (Nelson, J. I., Seiple, W. H., Kupersmith, M. J., & Carr, R. E. (1984). Lock-in Techniques for the Swept Stimulus Evoked Potential. Journal of Clinical Neurophysiology, 1(4), 409–436. doi:10.1097/00004691-198401040-00002) but could not find a mention of adaptation effects.

Other studies also measured adaptation effects on the CRF Heinrich and Bach, (IOVS, 2001), Ross and Speed (1991, Proc Roy Soc). And the hard-to-find Bach, Greenlee and Buhler (Clin Vis Sci, 1996) paper that modeled the effect within a contrast gain control framework.

> The Ross and Speed paper measures masking using SSVEP but the other adaptation measures are, as far as we can tell, psychophysical.  Heinrich and Bach are concerned with transient changes in SSVEP amplitude within a recording which may have an adaptation origin but which manifest as response increases over the first few seconds and seems a little mysterious (we often observe these as well). Bach, Greenlee and Buhler is, however, pretty unequivocal and we now include it. We like the suggestion that the  'increase' in response is attributable to a shift from a square wave response to a more sinusoidal waveform as the apparent contrast of the stimulus is decreased by adaptation-driven gain control. The response power which was distributed across odd harmonics in the original response can now be concentrated in the 1st harmonic after adaptation.

@@

 It’s good to have cited Regan and Regan paper that introduced the notion of fingerprints in the spectrum being tied to the nature of the non-linearity. Maybe expand their “fingerprint” approach a little more.

> We have expanded a little but are also conscious of limitations to this approach (see also the discussion re unique inputs and system order below). We think there might be some ways to relax these constraints a little by incorporating information about likely transducer functions at different stages (essentially fitting parameters rather than performing a full system identification) but this still requires many inputs.

Figure 4. Oversaturation occurs because the denominator overtakes the numerator at high contrast as it has two terms in it, not the single term in the numerator. Would be worth showing total power not saturating. Is this a general effect?

> We think this topic is probably worth a more detailed treatment than we can provide here. Broadly, much of the oversaturation effect can be modeled by an increase in saturation and therefore harmonic distortion - power is transferred from the first to higher harmonics as contrast increases and the output becomes more 'square-wave-like'. It seems likely that total power does not saturate (perceptually, there appears to be a monotonic increase in contrast for example and discrimination thresholds increase up to 100%) but the issue is complicated by the fact that the distribution of neurons sensitive to different contrast levels might not be uniform and that neurons might alter their response properties to signal with optimum sensitivity around a particular range. Our goal here is simply to show people new to the field how the CRF is approximately log-linear with contrast.

@@

Page 8 figure 5. The extrapolation to zero amplitude method should be attributed to Cambell and Maffei (J Physiol) 1970). Campbell and Kulikowski (J Physiol, 1972) made a detailed comparison to psychophysics.

> We have now included these references

Figure 5 is an oversimplification. As it doesn’t take into account how the additive noise mixes with the internal VEP response.  This was modeled in Norcia et al., (Vision Res, 1989). The additive noise adds a bias to the measurement, especially if the slope of the response function is shallow. The authors should try to connect the linear regression approach for threshold measurement to the sigmoidal non-linearity, as they are different models of the CRF. In practice the difference at the low contrast end is probably too small to matter and the saturation region is excluded.

> We have now note that it is important to consider noise for situations where the SNR is low or the slope is shallow and we expand on the point that you have to stick to the part of the CRF where the curve is more or less linear - as we showed explicitly in the figure. The code also includes an explicit 'noise baseline' that can be used to demonstrate changes in the intercept as a function of SNR.

@@

> Thank you for this explanation. We have now reworded the paper to make this clearer. We did take care to define what we mean by 'on-off' flicker in the preceding sentence but we now clarify the alternative 'appearence/disappearence' terminology. The point we were making was that odd harmonic responses (1F) are a feature of this this type of modulation because the responses to the appearance and disappearance of a pattern (or mean field) need not be completely symmetrical (See Zemon, et al 1988 for an explicit demonstration).

This type of stimulation does not separate contributions from “independent populations of on- and off-cells.”   There are other ssVEP techniques developed that do tap ON and OFF pathways selectively, which are not cited here. Contrast-reversal stimuli which involve symmetrical modulation of light and dark elements cancel odd harmonic frequency components in the response, however, both kinds of stimulus techniques mentioned here elicit only nonlinear responses.

> We have clearly not been sufficiently clear here. We do not think we suggested any form of 'separation' and we agree that a) there are infrequently-used paradigms that can target ON and OFF pathways more effectively (for example, 'sawtooth' temporal modulations that provide asymmetrical onset or offset transients) and b) responses from spatial features of differing polarity will tend to be combined at the scalp as long as the spatial frequency is high enough. We have expanded this section to attempt to make the general point clear.

@@

> We have included three earlier papers by Regan on this topic.

4. The statement “This normalization, achieved through a computation called ‘contrast gain control’” again neglects to cite the original work that defined “contrast gain control” and the authors ignored a nonlinear biophysical model proposed to explain this phenomenon in ssVEP contrast response functions (which was also applied in developmental work on infants and children).

> Contrast gain control is related to 'automatic gain control' - a electrical engineering concept that arose in the early 20th century. The concept of modulating the input (rather than the output) of the system became established in the 1930s and 40s (for example in WW2 radar systems). The specific idea of computing that modulation as a function of the local average signal power seems to have arisen around this time and was refined over the 50s and 60s (for example, in 'sidechain compression'). In the domain of vision, an early example of 'contrast gain control' (scaling the input - in this case luminance) is found in Sperling and Sondhi 1968.

The first formalization of the concept in the domain of spatial vision that included the critical concepts of an early, divisive gain control signal computed from some additional estimate of local contrast seems to appear in a seminal paper by Shapley and Victor, 1978. Application to contrast adaptation was demonstrated by Ohzawa, Sclar, Freeman, 1982, 1985] with a demonstration of instantaneous gain control in the human SSVEP by Bobak et al in 1988. A synthesis that included both adaptation and masking was proposed by Heeger 1992 while Foley 1994 show an application to human psychophysical data. We have now cited these more foundational papers. We would be happy to include the relevant references to the non-linear biophysical model used in adult and developmental work if the reviewer can point us to the appropriate literature.

@@

1. The description of Figure 3 and the nonlinear operations performed are unclear. They state in the text that Equation 2 (“hyperbolic ratio function”) is used as a nonlinear transducer, but the figure legend states that a squaring operator was used for frequency doubling. A squaring operation would only produce a sinusoid of double the frequency of the input. The waveform shown in the left panel appears to be a full-wave rectified signal, which would contain theoretically an infinite number of higher harmonics. The text also states that “This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity.” When was full-wave rectification performed?

> We agree that we went too fast here and the description of the signal processing was confusing. We now break out the two stages into (Figure 3) full wave rectification only (to model the contributions of on and off cells) - explicitly pointing out that this produces only even harmonics but does not alter the linearity of the CRF. We then show (Figure 4) the effect of passing this full wave rectified signal through a hyperbolic ratio function and illustrate the resulting saturation of the CRF (Figure 4). We think that breaking the model out in this way makes the contribution of each part clearer and allows the reader to play with the code for separate stages. The input contrast is squared in Figure 4 simply because the exponent we choose in the saturating nonlinearity is '2' but this is a parameter that is often fitted rather than fixed and the reader is encouraged to experiment with the exponent to examine its effect on the resulting amplitude spectrum.

While there are, technically, infinite even harmonics in both Figure 3 and Figure 4, they decrease with frequency and we plot only three for reasons of space.

@@

6. The authors use the term “frequency tagging” which has been used in the literature to refer to stimulating different regions of the stimulus field or fellow eyes with different sinusoidal frequencies and measuring the intermodulation (sum and difference) frequency responses that result from nonlinear interactions. They do not, however, cite the foundational work in this field.

> As far as we are aware, the term refers to stimulating some *aspect* of the visual scene (a position or a feature) - not necessarily just a spatial region. There need not be more than one region (it is interesting, for example, to frequency tag a single feature or region and measure the effect of attending towards or away from it) and frequency tagged components can exist in the same location (for example, see the work on contrast normalization described above where the 'probe' and 'mask' have different tags and we explicitly examine the intermodulation terms). In fact much of this paper is concerned with discussing precisely these intermodulation terms and their origins. However, intermodulation is also not a 'necessary' part of a frequency tagging experiment: beautiful early work by Morgan et al (PNAS 1996) uses spatial frequency tagging to examine attentional modulation purely in terms of the amplification or suppression of the F1 and F2 components.  The tag need not even be visible - there is a fashion at the moment for 'rapid invisible frequency tagging' (RIFT) using flicker close to or beyond the CFF limit. In this sense, steady state paradigms are almost by definition frequency tagging paradigms because at least one feature or position is 'tagged'. In terms of using multiple input frequencies to characterise interactions between visual features, we think that Regan and Regan may have had the most comprehensive and some of the earliest descriptions of the method and we have credited them extensively throughout the paper. We are also happy to consider additional works if the reviewer can suggest them.

@@

7. There is a large body of work that used ssVEPs to investigate contrast processing developmentally (infants through adulthood) and in various disease states, which is mostly missed in this review article.

> We touch on both these areas but including them in detail would require us to extend what is already becoming a very long paper.

8. Figures should be labeled.
> It is possible that our text compilation stage failed to auto-generate the figure labels - we apologise if this is the case. The current version seems to be working correctly (in that each Figure has a label saying 'Figure 1', 'Figure 2' etc which is referenced in the main text. )

Thus, the stated intention of this manuscript (“Here we describe how SSVEPs have been used to study visual contrast over the past 70 years”) fails to be met. Only select articles from a few sources are covered, and therefore, the richness and value of this technique is lost.

> The original document had well over 100 references and the current bibliography now contains 156 entries. Although we have tried to be fair, we have of necessity 'selected' from the literature and have doubtless omitted many important foundational works. We are happy to take suggestions if parts of the field have been misrepresented.

The original purpose of the document (now long eclipsed) was to serve as a convenient primer on the history of the technique and an interactive visualisation of key concepts implemented in Python and R using a novel 'executable paper' format. We think that getting this balance right means that we cannot include a comprehensive history of the entire field - in effect we must continue to be 'selective' and we hope that the relatively few papers we have cited here will serve as useful signposts to the richer literature.

### 2024-12-02 review.qmd
At the end. I think it needs checking..

  output=RMax*(inputContrast**expnt)/(inputContrast**expnt + maskContrast**expnt + c50**expnt)
  return output

@@

# Define some constants
nLevels=8
internalNoise=1
inputFrequency_Hz=np.array([5]) # We simulate a system with on input at many different contrasts
contrastRange=np.logspace(1,1.9,nLevels)/100 # Equally spaced log contrast between 1 and 10^1.5
duration_S=1 # All variables in this code have an _[units] where possible
digitisationRate_Hz=1000
freqCutoff_Hz=30

@@

initialSignal = np.zeros(len(contrastRange))
transducedSignal = np.zeros(len(contrastRange))
for contrastIndex,thisContrast in enumerate(contrastRange):
  inputModulation = np.sin(2 * np.pi * inputFrequency_Hz * support_S) * thisContrast + np.random.rand(len(support_S))*internalNoise

@@

Although its mathematical definition is straightforward, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background (see @fig-centresurround; note that the code used to produce all figures in this review is available in python and R at: https://github.com/wadelab/contrastReviewPaperVNS).

@@

To illustrate the effect of contrast gain control [@Heeger1992], we next include a second component (a 'mask') that contributes to the gain pool of the first (the 'target'). The mask stimulus will suppress the response at the target frequency, reducing its amplitude. The suppression is reciprocal - activity at the mask frequency is also reduced by the presence of the target, in a contrast-dependent manner [@Regan1988, @brownDevelopmentRivalryDichoptic1999, @Candy2001, @Busse2009]. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the transduced waveforms (left panel of @fig-twoinputs) and in the Fourier spectrum (middle panel of @fig-twoinputs). The two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988].

>>>>

  output=RMax*(inputContrast**expnt)/(inputContrast**expnt + maskContrast**expnt + c50**expnt)
  # Changing this equation can produce a different pattern of intermodulation terms in Figure 5.
  #output=RMax*(inputContrast**expnt + maskContrast**expnt)/(inputContrast**expnt + maskContrast**expnt + c50**expnt)
  return output

@@

# Define some constants
nLevels=8
spontaneousNoise=.1 # Changing the level of background noise can alter the slope of the regression line and therefore the estimate of the threshold

inputFrequency_Hz=np.array([5]) # We simulate a system with on input at many different contrasts
contrastRange=np.logspace(.5,1.9,nLevels)/100 # Equally spaced log contrast between 1 and 10^1.5
duration_S=1 # All variables in this code have an _[units] where possible
digitisationRate_Hz=1000
freqCutoff_Hz=30

@@

initialSignal = np.zeros(len(contrastRange))
transducedSignal = np.zeros(len(contrastRange))
for contrastIndex,thisContrast in enumerate(contrastRange):
  inputModulation = np.sin(2 * np.pi * inputFrequency_Hz * support_S) * thisContrast + np.random.rand(len(support_S))*spontaneousNoise

@@

Although its mathematical definition is straightforward, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@ohzawaContrastGainControl1982, @ohzawaContrastGainControl1985, @bobakCorticalContrastGain1988a, @Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background (see @fig-centresurround; note that the code used to produce all figures in this review is available in python and R at: https://github.com/wadelab/contrastReviewPaperVNS).

@@

To illustrate the effect of contrast gain control [@ohzawaContrastGainControl1982, @Heeger1992], we next include a second component (a 'mask') that contributes to the gain pool of the first (the 'target'). The mask stimulus will suppress the response at the target frequency, reducing its amplitude. The suppression is reciprocal - activity at the mask frequency is also reduced by the presence of the target, in a contrast-dependent manner [@Regan1988, @brownDevelopmentRivalryDichoptic1999, @Candy2001, @Busse2009]. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the transduced waveforms (left panel of @fig-twoinputs) and in the Fourier spectrum (middle panel of @fig-twoinputs). The two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988].
